Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-03-21 paper/paper.tex
thorough proofread 3 (beginning of section 5) * sorted and reworded common experiment setup * removed repeating pieces about replication * renamed and shortened a little bit the `stageswise` section

\documentclass[a4paper,10pt,onecolumn]{article}

@@

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that have been studied in prior research on complex-valued networks.

% common experiment settings
Every experiment is replicated several times to take random effects into account, e.g.
random network initialization prior to training, randomly ordered batches during SGD.
Unless specified otherwise, every experiment uses global gradient clipping within an
$\ell_2$-norm ball of radius $0.5$, runs SGD with ADAM optimizer with the base learning
rate of $10^{-3}$. We note that since our experiments are on classification tasks,
$\cplx$-valued networks ultimately have to output a real-valued logit score. Following
\citet{wolter_complex_2018} and \citet{trabelsi_deep_2017}, the class scores are taken
as real parts of the $\cplx$ output.

\subsection{Stagewise fitting and sparsification} % (fold)
\label{sub:fitting_and_sparsification}

The goal of this section to study the compression-performance trade-off of the proposed Bayesian
sparsification methods for $\cplx$-valued networks. To this end we trained each network in
three successive stages in every experiment: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,.
    \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.

The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. 
%
The benefit of the ``fine-tune'' stage appears to be strongly influenced by the compression
rate. We can observe in this and, more prominently, in MusicNet experiment (sec.~\ref{sub:musicnet})
that higher compression rate render the model more regularized, leaving less capacity for
overfitting.
% more ``holes'' Bayesian dropout ``poked'' makes the ``fine-tune'' serve as extended training.

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:variational_dropout}). Since
the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter is
considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise. Figure~\ref{fig:hist__and__threshold__tradeoff} is constructed
for $C \in \{\tfrac1{20}, \frac1{200}\}$ before running the ``fine-tune'' stage. From
\eqref{eq:elbo_with_coef}, the relative positions of the compression curves on this
plot, and the results of experiments below it can be concluded that $C$ has much more
substantial impact on the sparsity and performance, than the choice of $\tau$. Furthermore,
for $\tau > 0$ the performance quickly saturates while compression creeps downwards.

% on chosing the threshold
The threshold $\tau$ is picked so, that the remaining \textit{``non-zero''} parameters
are within $\delta$ relative tolerance of their mean with high probability under the
approximate posterior $q_\theta(W)$. This ensures that stochastic parameters are sufficiently
concentrated around $\mu$, which are used as stand-in values for network weights $W$,
when it runs in inference mode. Since $W$ is $\real$- or a proper $\cplx$-Gaussian, the
random variable $
  \tfrac{k \lvert W - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with $k=1$ ($\real$) or $2$ ($\cplx$).
%
For a generous tolerance $\delta = 50\%$ all values of $\log \alpha$ below $-2.5$ yield
at least $90\%$ chance of a non-zero $\real$/$\cplx$ parameter. We pick $\tau = -\tfrac12$
to exclude noisy parameters and encourage higher sparsity, at the same time being aware
that $q_\theta$ is merely an approximation. In comparison, $\tau = 3$ is commonly used as
the threshold \citep{molchanov_variational_2017,kingma_variational_2015}.

@@

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of floating point values needed to
store the model and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros, $n_\mathtt{par}$ is the total
number of floats. Thus each parameter is counted once in a $\real$-valued network, but twice
in a $\cplx$-network. Each model has compression limit, set by layers' biases, shift
and scaling from $\real$-and $\cplx$-batch normalization.

@@

In each experiment we vary the method and the value $C$, while keeping $\tau$ at $-\tfrac12$.
Every experiment is replicated five times to take into account the random effects from
initialization, batch sequence during SGD, and stochasticity of intermediate layers on
forward pass at ``sparsify'' stage.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=\columnwidth]{grad_log.pdf}
  \caption{$\tfrac{d f(\alpha)}{d \log{\alpha}}$ of the fit proposed by
  \citet{molchanov_variational_2017}, MC estimate of \eqref{eq:improper-kl-div-real},
  and the exact derivative using \eqref{eq:real-kl-div-deriv-log}.}
  \label{fig:molchanov-derivative-replica}
\end{figure}

@@

We recap the set-up of each additional experiment on MNIST-like datasets:
\begin{itemize}
  \item staged training, with weight and optimizer state transfer
  \item $40$, $75$, and $40$ epochs for ``dense'', ``sparsify'', and ``fine-tune'' stagers, respectively
  \item ADAM optimizer with learning rate $10^{-3}$ before the $10$-th, and $10^{-4}$
  afterwards
  \item global $\ell_2$-norm gradient clipping set at $\tfrac12$
  \item training batches of size $128$ from a fixed randomly subset ($10^4$ images) of the training split
  \item the multiplier of the Kullback-Leibler term in ELBO varies over $
    C \in \{
      \tfrac32 2^{-\tfrac{k}2} \colon k=2, \cdots, 38
    \}
  $
  \item the sparsification threshold $\tau$ is set to $-\tfrac12$
\end{itemize}
Each experiment has been performed $5$ times to obtain a small sample of performance-compression
values for each setting of $C$. By reevaluating each setting several times we can account
for randomness in initialization of the uncompressed network at ``dense'', stochasticity of
variational dropout layers, randomness of batch sequences during SDG, and non-determinism
of computations on GPUs.

>>>>

\documentclass[a4paper,10pt,twocolumn]{article}

@@

We study the compression-performance trade-off of the proposed Bayesian sparsification
methods for $\cplx$-valued networks presented above on the datasets, that have been
studied in prior research on complex-valued networks.

% setup and common experiment settings
For each dataset we conduct a set of experiments in which we try out all combinations of
model kinds ($\real$ or $\cplx$), sparsification methods (VD or ARD) and input features.
Where reasonable, we also study the effects of halving ($\cplx$) or doubling ($\real$)
of the size of intermediate layers in each network \citep{monning_evaluation_2018}.
%
Each network within every experiment passes through three successive stages of training:
pre-training, sparsification, and fine-tuning (sec.~\ref{sub:fitting_and_sparsification}).
%
Each experiment is replicated \textit{five} times to take into account random initialization,
random order of the mini-batch during SGD, stochastic outputs of the intermediate layers,
and non-determinism of single-precision computations on GPUs.

Unless specified otherwise, every experiment uses global gradient clipping within an
$\ell_2$-norm ball of radius $0.5$, employs SGD with ADAM optimizer with the base learning
rate of $10^{-3}$. We note that since our experiments are on classification tasks,
$\cplx$-valued networks ultimately have to output a real-valued logit score. Following
\citet{wolter_complex_2018} and \citet{trabelsi_deep_2017}, the class scores are taken
as real parts of the $\cplx$ output.

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of IEEE754 floating point values
needed to store the model and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros, $n_\mathtt{par}$ is the total
number of floats. In particular, each parameter is counted once in a $\real$-valued network,
but twice in a $\cplx$-network. Each model has a compression limit, determined by its
layers' biases, shift and scaling from $\real$-and $\cplx$-batch normalization.

\subsection{Stagewise training} % (fold)
\label{sub:fitting_and_sparsification}

% The training consists of three successive stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture, and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.

During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$ counterpart.
We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo} to study the attainable sparsity levels:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:elbo}'}
\end{equation}
By varying $C$ we adjust the importance of the sparsifying prior in learning $q_\theta(W)$,
meaning that higher value yields more irrelevant parameters during in inference and makes
layers sparser upon termination of this stage.

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:variational_dropout}).
Since the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter
is considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise.

The \textit{``fine-tune''} stage reverts the network back to non-Bayesian layers, applies
sparsity masks to its parameters and proceeds with training the kept parameters as during
the ``dense'' stage. Unless specified otherwise the network is initialized to the mode
of the learnt $q_\theta(W)$.

% on chosing the threshold
The threshold $\tau$ is picked so, that the remaining non-zero parameters are within
$\delta$ relative tolerance of their mode with high probability under the approximate
posterior. Since $q_\theta$ is fully factorized $\real$- or a proper $\cplx$-Gaussian,
the random variable $
  \tfrac{k \lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with $k=1$ ($\real$) or $2$ ($\cplx$).
%
For a generous tolerance $\delta = 50\%$ all values of $\log \alpha$ below $-2.5$ yield
at least $90\%$ chance of a non-zero $\real$/$\cplx$ parameter. We pick $\tau = -\tfrac12$
to retain parameters sufficiently concentrated around their mode, and encourage higher
sparsity, at the same time being aware that $q_\theta$ is merely an approximation.
In comparison, $\tau = 3$ is commonly used as the threshold
\citep{molchanov_variational_2017,kingma_variational_2015}.

@@

% conlcusions
The impact of the ``fine-tune'' stage appears to be strongly influenced by the compression
rate. We can observe in this and, more prominently, in MusicNet experiment (sec.~\ref{sub:musicnet})
that higher compression rate render the model more regularized, leaving less capacity for
overfitting.
% more ``holes'' Bayesian dropout ``poked'' makes the ``fine-tune'' serve as extended training.

% effects of extra training epochs and the impact of tau
Figure~\ref{fig:hist__and__threshold__tradeoff} is constructed
for $C \in \{\tfrac1{20}, \frac1{200}\}$ before running the ``fine-tune'' stage. From
\eqref{eq:elbo_with_coef}, the relative positions of the compression curves on this
plot, and the results of experiments below it can be concluded that $C$ has much more
substantial impact on the sparsity and performance, than the choice of $\tau$. Furthermore,
for $\tau > 0$ the performance quickly saturates while compression creeps downwards.

@@

@@

In each experiment we vary the method and the value $C$, while keeping $\tau$ at $-\tfrac12$.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=\columnwidth]{grad_log.pdf}
  \caption{$\tfrac{d f(\alpha)}{d \log{\alpha}}$ of the fit proposed by
  \citet{molchanov_variational_2017}, MC estimate of \eqref{eq:improper-kl-div-real},
  and the exact derivative using \eqref{eq:real-kl-div-deriv-log}.}
  \label{fig:molchanov-derivative-replica}
\end{figure}

@@

We recap the set-up of each additional experiment on MNIST-like datasets:
\begin{itemize}
  \item staged training, with weight and optimizer state transfer
  \item $40$, $75$, and $40$ epochs for ``dense'', ``sparsify'', and ``fine-tune'' stagers, respectively
  \item ADAM optimizer with learning rate $10^{-3}$ before the $10$-th, and $10^{-4}$
  afterwards
  \item global $\ell_2$-norm gradient clipping set at $\tfrac12$
  \item training batches of size $128$ from a fixed randomly subset ($10^4$ images) of the training split
  \item the multiplier of the Kullback-Leibler term in ELBO varies over $
    C \in \{
      \tfrac32 2^{-\tfrac{k}2} \colon k=2, \cdots, 38
    \}
  $
  \item the sparsification threshold $\tau$ is set to $-\tfrac12$
\end{itemize}
Each experiment has been performed $5$ times to obtain a small sample of performance-compression
values for each setting of $C$.
